---
title: "Seattle Data Analysis"
output:
  html_document:
    df_print: paged
---

<style>
body {
text-align: justify}
</style>

### Introduction

The Seattle Public Library dataset contains information on the checkout history of books related to Harry Potter written by J.K. Rowling. The dataset includes 12 columns, such as UsageClass, CheckoutType, MaterialType, CheckoutYear, CheckoutMonth, Checkouts, Title, ISBN, Creator, Subjects, Publisher, and PublicationYear. The UsageClass column contains two unique values: Digital and Physical, and there are two CheckoutType values: Horizon and Drive. The MaterialType column has two values: BOOK and EBOOK, and the CheckoutYear ranges from 2012 to 2022.

In this analysis, we will focus on the trends of the Harry Potter books, both digital and physical, and compare the checkout patterns of the two types. Specifically, we will analyze the checkout trends over the years, looking at the Checkouts column, and examine the most popular publisher based on the number of checkouts. We will also investigate the preferred checkout type, Horizon or Drive, and see how average number of checkouts varry over the years. We chose to analyze these trends because it can provide valuable insights into the popularity of Harry Potter books over the years, and the reading preferences of library users in Seattle.

### Summary Statistics

```{r}
#reading data from csv file
data <- read.csv("HP_Checkouts_by_Title.csv")
head(data)
```


```{r, warning=FALSE, message=FALSE}
#average checkouts by check out type
library(dplyr)
data%>%
  group_by(CheckoutType)%>%
  summarise(avg_checkouts = mean(Checkouts))
```

From the above output, the average checkouts for Horizon is 14.08 which is much lower than average checkouts for check out type over drive.


```{r}
#Year with most checkouts
data%>%
  group_by(CheckoutYear)%>%
  summarise(Checkouts = sum(Checkouts))%>%
  arrange(desc(Checkouts))
```


From the above output, it can be observed that 2017 has highest number of checkouts (5353) followed by 2018 with checkouts (4971) and the year 2020 has lowest number checkouts (1171) followed by year 2021 (1682).

```{r}
#Most/least number of checkout by Month
data%>%
  group_by(CheckoutMonth)%>%
  summarise(Checkouts = sum(Checkouts))%>%
  arrange(desc(Checkouts))%>%
  arrange(desc(CheckoutMonth))
```

From the above output, December has highest number of checkouts followed by November. January has lowest number of Checkouts followed by February.


```{r}
#average checkouts by Material Type
data%>%
  group_by(MaterialType)%>%
  summarise(avg_checkout = mean(Checkouts))
```

From the above output, the book material type highest average checkouts (14), which is much lower as compare to the ebook material type average checkouts (8.43).


```{r}
#summarizing  the Checkouts by CheckoutType
data%>%
  group_by(Publisher)%>%
  summarise(avg_checkouts = mean(Checkouts))%>%
  arrange(desc(avg_checkouts))
```

From the above output, it can be seen that Arthur A. Levine Books has highest average checkouts (26.23), followed by Scholastic (25.63),, and Arthur A. Levine Books, an imprint of Scholastic Inc., (20.63). The Huang guan wen hua chu ban you xian gong si has average checkout value of 1.078947 followed by Nhà xuất bản Trẻ with average checkouts (1.09).

### Data
The data was collected and published by the Seattle Public Library. The dataset contains information on the checkout history of books related to Harry Potter written by J.K. Rowling. The parameters of the data include the checkout date, checkout type (Horizon or Drive), material type (book or ebook), usage class (digital or physical), title, ISBN, creator, subjects, publisher, and publication year. The data spans from 2012 to 2022. The data was collected automatically by the library's checkout system whenever a user checked out a Harry Potter book. The data was collected to track the usage and popularity of Harry Potter books in the Seattle Public Library, to inform collection management decisions, and to provide insights into reading preferences. When working with this data, it is important to consider ethical questions related to user privacy and data protection. The dataset includes information about the titles that users have checked out, which could potentially reveal sensitive information about their interests or beliefs. To address this, the library likely took steps to anonymize the data, such as removing user identifiers or aggregating data at the level of the book rather than the individual user. There are several limitations or problems with this data that should be considered. Firstly, the dataset only includes information on Harry Potter books written by J.K. Rowling, so it may not be representative of the popularity of other books or authors in the library's collection. Additionally, the dataset is limited to books checked out from the Seattle Public Library, which may not be representative of the reading habits of the broader population. The data is also limited to the time period from 2012 to 2022, so it does not provide a comprehensive view of the popularity of Harry Potter books over time. Finally, there may be biases in the data collection process, such as differences in checkout practices between different library branches or changes in the checkout system over time, which could affect the accuracy of the data. Overall, these limitations should be taken into account when interpreting the results of any analysis conducted using this data.



### Data Visualization

#### Number of Checkouts by Material Type for each Year

For the first plot, the total number of checkouts will be visualized for each year with different lines for material type.

```{r, warning=FALSE, message=FALSE}
# Create a new column for year
library(ggplot2)
# Aggregate data by MaterialType and Year
data_agg <- aggregate(Checkouts ~ MaterialType + CheckoutYear, data, sum)

# Create line graph
ggplot(data_agg, aes(x = CheckoutYear, y = Checkouts, color = MaterialType)) +
  geom_line() +
  labs(title = "Number of Books Checked Out by Material Type",
       x = "Year",
       y = "Number of Checkouts",
       color = "Material Type") +
  scale_x_continuous(breaks = seq(2012, 2022, 1)) 
```


It is clear from the above plot that the total number of checkouts for the book material type initially increased from the years 2012 to 2013, then decreased in the following year, 2014. It then had a rising tendency, reaching its peak value in the year 2017, and then dropping to a lower value in the year 2020. The years 2021 and 2022 saw a subsequent uptick, nevertheless. The lowest total checkout for ebook material type was recorded in the year 2012, while the greatest total checkout was recorded in the year 2017. On the other hand, the overall trend for total checkouts of ebook material type showed continuous growth. In particular, the plot leaves out crucial details about ebook checkouts. The book material type has higher number of checkouts as compare to the ebook material type.

For the second plot, the total number of checkouts for some of the publisher will be visualized for each year.

#### Number of Checkouts by Publisher for each Year


```{r}
#filtering data for publisher Arthur A.Levin Books, Pottermore, Scholastic, Salamandra and Seizansha.
data2 <- data%>%
  filter(Publisher %in% c("Arthur A. Levine Books,", "Pottermore",
                          "Scholastic,", "Salamandra,", "Seizansha,"))
# Aggregate data by MaterialType and Year
data_agg <- aggregate(Checkouts ~ Publisher + CheckoutYear, data2, sum)

# Create line graph
ggplot(data_agg, aes(x = CheckoutYear, y = Checkouts, color = Publisher)) +
  geom_line() +
  labs(title = "Number of Books Checked Out by Different publisher",
       x = "Year",
       y = "Number of Checkouts",
       color = "Material Type") +
  scale_x_continuous(breaks = seq(2012, 2022, 1)) 
```


From the above plot, it can be seen that for publisher Arthur A. Levine Books, has highest number of total checkouts. For Arthur A. Levine Books, the number of checkout increased from 2012 to 2013, and then again decreased in 2014 and increasing again and reached it to maximum value in 2018 and then again started to decreased and reached to lowest checkouts in 2020 and then again increased in 2021 and 2022. The Pottermore publisher has second highest number of checkouts overall. For Pottermore, the overall trend is increasing from 2012 to 2018 and reached to maximum value in 2017. For Salmandra, the number of checkouts from 2012 to 2015 is almost same before it started to increase and then again decreased. For Scholastic publisher, the number of checkouts from 2012 to 2014 is almost same and then it started to increase and reached to maximum value in 2019 and then started to decrease again. For Seizansha, the number of checkouts are almost same for 2012 to 2022.



#### Distribution of Checkout by Year


For this plot, I will use a bar plot to visualize the distribution of number of checkouts for each year. The reason for this plot is to identify for which year the number of checkouts are highest and for which year the number of checkouts are lowest.
```{r}
#distribution of checkout by year
data$CheckoutYear <- as.factor(data$CheckoutYear)
data2 <- data[data$Checkouts < 300,]
data2%>%
  group_by(CheckoutYear)%>%
  summarise(Checkouts = sum(Checkouts))%>%
  ggplot(aes(x = reorder(CheckoutYear, -Checkouts), y = Checkouts, fill = CheckoutYear)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Checkouts by Year", x = "Year") +
  theme(legend.position = "none")
```

From the above plot, it can be seen that year 2017 has highest number of checkouts followed 2016 and 2018. The plot also shows that the year 2020 has lowest number of checkouts followed by 2021 and 2022. The year 2015, 2013, 2019, and 2014 has minor difference in the number of checkouts.

